The Precambrian paleogeography of Laurentia

### 2020-10-02 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits to Australia section and a commit of a compiled PDF

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}. The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interprets this relationship as persistent from 1.7 to 0.7 Ga. The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny. The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a}. This configuration back to ca. 1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt. Isa dikes pole from North Australia and is consistent with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca. 1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018a}) with further motion by ca. 1070 Ma. The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia. Such correlations have led to a number of different proposed connections that place Australia + East Antarctica along the western margin of Laurentia at different times in the Proterozoic. Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to the be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwest Laurentia. The ca. 1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.

@@

There is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there is also an obducted ophiolite as well as eclogites preserved in the Trans-Hudson orogen \citep{Weller2017a}. These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.

>>>>

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}. The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interprets this relationship as persistent from 1.7 to 0.7 Ga. The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny. The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia. This configuration back to ca. 1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt. Isa dikes pole from North Australia and with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca. 1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018a}) with further motion by ca. 1070 Ma. The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia \citep{Moores1991a}. The extent of Antarctic lithosphere that was conjoined with Antarctica prior to the assembly of Gondwana is uncertain, but there are strong connections between the Gawler province of the South Australia craton and the Terre Ad\'elie province of Antarctica that is commonly interpreted to extended to the Miller ranges of the Trans-Antarctic Mountains as the Mawson craton \citep{Payne2009a}.  Separation between these Antarctic Provinces and Australia was largely accomplished in the current Cenozoic Era with the opening of the Tasmanian seaway. Correlations with Laurentia have led to a number of different proposed reconstructions of Australia + East Antarctica along the western margin of Laurentia at different times in the Proterozoic.

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwest Laurentia. The ca. 1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. If the North Australian craton was continous with the South Australia his interpretation would have the ca. 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca. 1.35 Ga East Kootenay orogeny. Another challenge with interpreting data from North Australia A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.

@@

Even without considering other continents, there is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there is also an obducted ophiolite as well as eclogites preserved in the Trans-Hudson orogen \citep{Weller2017a}. These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.
